Canonical rings of rational divisors on projective space and Hirzebruch surfaces

### 2015-07-16 surfaces-write-up/surface-write-up.tex
section 5.2 edits

%%%%%%%%%%%%%%%%%%%%%%% amsthm theorem styles %%%%%%%%%%%%%%%%%%%%%%%

@@

\title{Canonical rings of $\bq$-divisors on projective space and Hirzebruch surfaces}

@@

Another closely related topic is the Hassett-Keel program 
initiated in \todo{was it actually initiated here? check vzb.} \cite{hassett:classical-and-minimal-models}, which aims to describe log canonical models of the form
\begin{align*}
	\overline {\mathscr M}_g(\alpha) := \bigoplus_{d \geq 0}H^0 \left( \overline {\mathscr M}_g, \lfloor d K_{\overline{\mathscr M}_g} + \alpha\delta \rfloor  \right) 
\end{align*}
in terms of certain moduli spaces. $\bq$-divisors on surfaces also naturally appear when considering the canonical ring of surfaces of Kodaira dimension 1. A surface has Kodaira dimension 1 if and only if it is an elliptic surface, and their canonical ring are most naturally described in the setting of $\bq$-divisors. \todo{Add references}

@@

\ssec{Notation}
\begin{convention}
Notate a (Weil) $\bq$-divisor
\begin{align*}
	D = \sum_{i=1}^{n}\alpha_i D_i \in \di X \otimes \bq.
\end{align*}

\noindent
where $\alpha_i \in \bq$ and $D_i \in \di X$ is an irreducible divisor.
\end{convention}

@@

In this subsection, we restrict attention to the case of effective fractional divisors $D \in \bq \otimes \di \bp^m$.

\begin{defn}
\label{defn:vec-sum}
For $\vec{v} \in \br^n$, denote $\deg \vec{v} := \sum_{i = 0}
^n v_i$.
Let 
\begin{align*}
	\mss_i := \left \{\vec{v} \in \bz_{\geq 0}^{m + 1} \; : \;
\deg \vec v = c_i \right\}.	
\end{align*}
\end{defn}

\begin{defn}
\label{defn:vec-order}
Let $\vec{v}, \vec{w} \in \br^n$. Let $i \in \{1,\ldots, n\}$
be the biggest index such that $v_i$ is nonzero
and $j \in \{1,\ldots, n\}$ be the smallest index such that $w_j$ is
nonzero. Define $\vec{v} \prec \vec{w}$ if $i \leq j$.
\end{defn}

\begin{thm}
\label{thm:proj-one-point}
Let $\bp^m \cong \proj \bk [x_0, \ldots, x_m].$ Let $D = \alpha H$, with $\alpha = \frac{p}{q} \in \bq$
and $H := V(x_k)$ a hyperplane of $\bp^m$.
Let
\[
	0 = \frac{c_0}{d_0} <
	\frac{c_1}{d_1} < \ldots < \frac{c_r}{d_r} = \frac{p}{q}
\]

\noindent
be the best lower approximations of $\alpha$. Then, the
canonical ring
\[
	R_D := \bigoplus_{d \geq 0} H^0(\bp^m, \lfloor dD \rfloor)
\]

\noindent
has a minimal presentation consisting of the $\sum_{i = 0}^{r}
{{m + c_i - 1} \choose {c_i}}$ generators $F_i^{\vec{v}} := \frac{u^{d_i}
x^{\vec{v}}}{x_k^{c_i}}$ where $1 \leq i \leq r$, $\vec{v} \in \bz_{\geq 0}^{m + 1}$
with $\deg \vec v = c_i$. Furthermore, it has
relations of the following two types:
\begin{enumerate}
	\item For each $(i, j)$ with $j \geq i + 2$ and each $\vec{v} \in \mss_i,
\vec{w} \in \mss_j$, there is a relation of the form either
\begin{flalign*}
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\vec{y} \in \mss_{h_{i, j}}} (F_{h_{i, j}}^{\vec{y}})
	^{g_{\vec{y}}} &(i < h_{i, j} < j) \\
	\text{or} \\
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\substack{\vec{y} \in \mss_{h_{i, j}} \\ \vec{z} \in
	\mss_{h_{i, j} + 1}}} (F_{h_{i, j}}^{\vec{y}})^{g_{\vec{y}}} \;
	(F_{k_{i, j} + 1}^{\vec{z}})^{g'_{\vec{z}}}
	&(i < h_{i, j} < h_{i, j} + 1 < j).
\end{flalign*}
	\item For each $(i, j)$ with
$j = i$ or $j = i + 1$ and each $\vec{v} \in \mss_i, \vec{w} \in
\mss_j$ with $\vec{v} \not\prec \vec{w}$ (see Definition
~\ref{defn:vec-order}) there is a relation of the form
\begin{align*}
	&L_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- F_i^{\vec{y}} F_j^{\vec{z}} \\
\end{align*}

\end{enumerate}

@@

\section{Canonical Rings on Hirzebruch surfaces}
Let $D=\sum_{i=1}^n \alpha_i D_i \in \mathbb{Q} \otimes \di(\hirz_
k).$ such
that $D_1, \ldots, D_4$ are distinct divisors with bi-degrees $(1,0)
, (1,0), (0,1)$, and $(0,1)$ respectively. Let $D_i = V(f_i)$ where
$f_i \in \sco(\bida_i, \bidb_i)$.  Suppose $f_1$ and $f_2$ are independent linear
polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent
linear polynomials in $y_0$ and $y_1$.

@@

\begin{proof}
Let $(\bida_i, \bidb_i)$ be the bi-degrees of $f_i$. Notice that if $\gamma
\in (R_D)_d$, then 
\[
	g = \sum_{i = 1}^n {f_i}^{c_i}
\]

\noindent
where each $c_i \ge - \alpha_i c_i$, $\sum_{i=1}^n c_i \bida_i = 0$,
and $\sum_{i=1}^n c_i \bidb_i = 0$. Then we can view $g$ inside of the
lattice $S \subseteq \mathbb{R}^{n + 1}$ given by ordered elements of
the form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge - d \alpha_i$
for all $i$, $\sum_{i=1}^n c_i \bida_i = 0$, and $\sum_{i = 1}^n c_i \bidb_i =
 0$. From here, determining a spanning set for $(R_D)_d$ reduces to
finding the extremal rays of $S$ by Proposition ~\ref{prop:cone-generation}.

To do this, we extend the method of O'Dorney \cite[Theorem 8]{dorney:canonical}. 
We first view the cone inside of $\mathbb{R}^{n+1}$ given by elements of the
form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge -d \alpha_i$ for
all $i$ and $\sum_{i = 1}^n c_i (\bida_i + \bidb_i) = 0$. By O'Dorney, this
has extremal rays given by

@@

Let $E$ be the set of $\epsilon_i$ and $\epsilon_{i,j}$ given 
above, satisfying the respective conditions.
The lattice points corresponding to elements of $E$ satisfies the conditions of Proposition ~\ref{prop:cone-generation}, so by applying that proposition, $R_D$ is generated in 
degree less than
\[
	\sigma = \sum_{i\in \Te(D)} \gcd(\bida_i, \bidb_i)\ell_i + \sum_{\substack{
	i \in \Tp(D) \\	j \in \Tm(D)}} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\end{proof}

Let $x_1, \ldots x_r$ be the generators of $R_D$ in degree less than $\sigma$ (as given by Lemma ~\ref{lem:hirz-generators}), and let $\phi$ be the surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through 
\[
	\Sigma =  \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d
	\alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i.
	\rangle 
\]
into
\[
\begin{tikzcd}
	\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
	x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]

Applying Lemma ~\ref{lem:bound-ker-chi}, we can bound the degrees of generation of $\ker(\chi)$ below
$2 \sigma$.
Finally, we calculate the degree of generation of $\psi$:

>>>>

%%%%%%%%%%%%%%%%%%%%%%% amsthm theorem styles %%%%%%%%
%%%%%%%%%%%%%%%

@@

\title{Canonical rings of rational divisors on projective space and Hirzebruch surfaces}

@@

Another closely related topic is the Hassett-Keel program 
initiated in \todo{was it actually initiated here? check vzb.} \cite{hassett:classical-and-minimal-models}, which aims to describe log canonical models of the form
\begin{align*}
	\overline {\mathscr M}_g(\alpha) := \bigoplus_{d \geq 0}H^0 \left( \overline {\mathscr M}_g, \lfloor d K_{\overline{\mathscr M}_g} + \alpha\delta \rfloor  \right) 
\end{align*}
in terms of certain moduli spaces. Fractional divisors on surfaces also naturally appear when considering the canonical ring of surfaces of Kodaira dimension 1. A surface has Kodaira dimension 1 if and only if it is an elliptic surface, and their canonical ring are most naturally described in the setting of fractional divisors. \todo{Add references}

@@

\ssec{Notation}
\begin{convention}
Notate a fractional (Weil) divisor
\begin{align*}
	D = \sum_{i=1}^{n}\alpha_i D_i \in \di X \otimes \bq.
\end{align*}
where $\alpha_i \in \bq$ and $D_i \in \di X$ is an irreducible divisor.
\end{convention}

@@

In this subsection, we restrict attention to the case of effective divisors.

\begin{defn}
\label{defn:vec-sum}
For $\vec{v} \in \br^n$, denote $\deg \vec{v} := \sum_{i = 0}
^n v_i$ (i.e. the taxicab metric).
Let $\mss_i := \{\vec{v} \in \bz_{\geq 0}^{m + 1} \; : \;
\sum_{i = 0} ^{m + 1} v_i = c_i\}$ be the set of points in
$\bz_{\geq 0}^{m + 1}$ with sum of components equal to $c_i$.
\end{defn}

\begin{defn}
\label{defn:vec-order}
Let $\vec{v}, \vec{w} \in \br^n$ be two vectors. Let $i$
be the right-most coordinate such that $v_i$ is nonzero
and $j$ be the left-most coordinate such that $w_j$ is
nonzero. Define $\vec{v} \prec \vec{w}$ if $i \leq j$.
\end{defn}

\begin{thm}
\label{thm:proj-one-point}
Let $X = \bp^m$ with $\{x_0, \ldots, x_m\}$ a coordinate system for
$\bp^m$. Let $D = \alpha H$, with $\alpha = \frac{p}{q} \in \bq$
and $H = V(x_k)$ a hyperplane of $\bp^m$, be a divisor on $\bp^m$.
Let
\[
	0 = \frac{c_0}{d_0} <
	\frac{c_1}{d_1} < \ldots < \frac{c_r}{d_r} = \frac{p}{q}
\]

\noindent
be the best lower approximations of $\alpha$. Then the
canonical ring
\[
	R_D := \bigoplus_{d \geq 0} H^0(\bp^m, \lfloor dD \rfloor)
\]

\noindent
has a minimal presentation consisting of the $\sum_{i = 0}^{r}
{{m + c_i - 1} \choose {c_i}}$ generators $F_i^{\vec{v}} := \frac{u^{d_i}
x^{\vec{v}}}{x_k^{c_i}}$ where $\vec{v} \in \bz_{\geq 0}^{m + 1}$
such that $\sum_{j = 0}^{m} v_j = c_i$. Furthermore, it has
relations of the following two types.

For each $(i, j)$ with $j \geq i + 2$ and each $\vec{v} \in \mss_i,
\vec{w} \in \mss_j$, there is a relation of the form either
\begin{flalign*}
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\vec{y} \in \mss_{h_{i, j}}} (F_{h_{i, j}}^{\vec{y}})
	^{g_{\vec{y}}} &(i < h_{i, j} < j) \\
	\text{or} \\
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\substack{\vec{y} \in \mss_{h_{i, j}} \\ \vec{z} \in
	\mss_{h_{i, j} + 1}}} (F_{h_{i, j}}^{\vec{y}})^{g_{\vec{y}}} \;
	(F_{k_{i, j} + 1}^{\vec{z}})^{g'_{\vec{z}}}
	&(i < h_{i, j} < h_{i, j} + 1 < j).
\end{flalign*}

Furthermore, there are the relations for each $(i, j)$ with
$j = i$ or $j = i + 1$ and each $\vec{v} \in \mss_i, \vec{w} \in
\mss_j$ with $\vec{v} \not\prec \vec{w}$ (see Definition
~\ref{defn:vec-order}) of the form
\begin{align*}
	&L_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- F_i^{\vec{y}} F_j^{\vec{z}} \\
\end{align*}

@@

\section{Canonical Rings on Hirzebruch surfaces}
Let $D=\sum_{i=1}^n \alpha_i D_i \in \mathbb{Q} \otimes \di(\hirz_
k).$ such
that $D_1, \ldots, D_4$ are distinct divisors with bi-degrees $(1,0)
, (1,0), (0,1)$, and $(0,1)$ respectively. Let $D_i = V(f_i)$ where
$f_i \in \sco(\bida_i, \bidb_i)$.  
Then, $f_1$ and $f_2$ are independent linear
polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent
linear polynomials in $y_0$ and $y_1$. \todo{Talk a little about Hirzebruch surfaces before this paragraph; especially, describe what it's sections look like.}

@@

\begin{proof}
%Let $(\bida_i, \bidb_i)$ be the bi-degrees of $f_i$. 
Notice that if $\gamma
\in (R_D)_d$, then 
\[
	g = \sum_{i = 1}^n {f_i}^{c_i}
\]

\noindent
where each $c_i \ge - \alpha_i c_i$, $\sum_{i=1}^n c_i \bida_i = 0$,
and $\sum_{i=1}^n c_i \bidb_i = 0$. Then we can any view $g$ inside the
lattice $S \subseteq \mathbb{R}^{n + 1}$ given by ordered elements of
the form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge - d \alpha_i$
for all $i$, $\sum_{i=1}^n c_i \bida_i = 0$, and $\sum_{i = 1}^n c_i \bidb_i =
 0$. From here, determining a spanning set for $(R_D)_d$ reduces to
finding the extremal rays of $S$ by Proposition ~\ref{prop:cone-generation}.

To do this, we extend the method of O'Dorney \cite[Theorem 8]{dorney:canonical}. 
We first consider the cone inside of $\mathbb{R}^{n+1}$ given by elements of the
form $(d, r_1, \ldots, r_n)$ satisfying $r_i \ge -d \alpha_i$ for
all $i$ and $\sum_{i = 1}^n r_i (\bida_i + \bidb_i) = 0$.  This
has extremal rays given by

@@

Let $E$ be the set of $\epsilon_i$ and $\epsilon_{i,j}$ given 
above, satisfying the respective conditions.
The lattice points corresponding to elements of $E$ satisfies the conditions of Proposition ~\ref{prop:cone-generation}, so by applying that proposition, $R_D$ is generated in 
degrees less than
\[
	\sigma = \sum_{i\in \Te(D)} \gcd(\bida_i, \bidb_i)\ell_i + \sum_{\substack{
	i \in \Tp(D) \\	j \in \Tm(D)}} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\end{proof}

Let $x_1, \ldots x_r$ be the generators of $R_D$ in degrees less than $\sigma$ (as given by Lemma ~\ref{lem:hirz-generators}), and let $\phi$ be the surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through 
\[
	\Sigma =  \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d
	\alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i.
	\rangle 
\]
into
\[
\begin{tikzcd}
	\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
	x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]

Applying Lemma ~\ref{lem:bound-ker-chi}, we can bound the degree of generation of $\ker(\chi)$ below
$2 \sigma$.
Finally, we calculate the degree of generation of $\psi$:
